---
title: Health and Economic Impacts from Severe Weather Events Analyzed from NOAA Storm
  Database
output:
  html_document:
    df_print: paged
---
```{r, echo=FALSE, message=FALSE}
library(tidyverse)

```

## Synopsis
In this report, we analyze a collection of storm data from a NOAA database to gain some insight into the most deleterious features of storms and severe weather events that have the most economic and human health impacts.

## Data Processing
Data is provided in CSV format with the first row containing column headers comprising aroughly 536MB of data. `tidyverse` library is used extensively throughout this analysis. For data processing, `read_csv` from this set of packages is used to read in the data.

```{r}
ds.raw <- read_csv('data/repdata_data_StormData.csv')
head(ds.raw)
print(dim(ds.raw))
```

```{r}
ds.casualty.summary <- group_by(ds.raw, EVTYPE) %>% 
                       summarize(TOT.FATALITIES = sum(FATALITIES), 
                                 TOT.INJURIES = sum(INJURIES))
```

